---
title: "Insuring Unemployment Out of Your Own Pension"
subtitle: "Causal Evidence from Mexico's *Retiro Parcial por Desempleo*"
thanks: ""
author: Esteban Degetau
email: estebandegetau@gmail.com
institute:  |
  Instituto Mexicano del Seguro Social, Barcelona School of Economics
date: 2026-08-14
date-format: long
# The venue printed alongside the date is \rpdvenue in slides/beamer-rpd.tex. It cannot
# live here: the title page renders only title/subtitle/author/institute/date, so a
# `venue:` key would be dropped without a warning -- as `email:` above already is.
bibliography: ../references.bib
# csl: ../amafore/apa.csl
format:
  beamer:
    # xelatex, not the lualatex default: under LuaTeX the beamer template loads
    # selnolig whenever `lang:` is set, selnolig is absent from the image, and
    # tlmgr's auto-install then dies on fmtutil-sys (the container user is uid
    # 1000 and /usr/local/texlive is root-owned). Same fix as iab/_metadata.yml.
    # For the same reason `lang:` is deliberately not set here.
    pdf-engine: xelatex
    aspectratio: 169
    # `t` top-aligns frame content; without it beamer centres it vertically and a
    # table ends up floating in the middle of the slide. `notheorems` is Quarto's
    # own default, repeated here because supplying classoption replaces it.
    classoption: [t, notheorems]
    fontsize: 11pt
    # NOT `mainfont:`. Libertinus is invisible to fontconfig -- it lives only in
    # the texmf tree -- so \setmainfont cannot find it, and setting it would also
    # switch the whole deck to serif. See slides/beamer-rpd.tex.
    fontfamily: libertinus
    section-titles: true
    slide-level: 2
    fig-align: center
    keep-tex: true
    latex-clean: false
    include-in-header: beamer-rpd.tex
execute:
  echo: false
  warning: false
  message: false
  freeze: false
  # The PDFs/PNGs in _figuras/ are the cache; a second knitr cache on top of them
  # only adds a way for the two to disagree.
  cache: false
---

```{r}
#| label: setup
#| include: false

pacman::p_load(targets, here, knitr, kableExtra, dplyr, tidyr, purrr, stringr, scales)
here::i_am("slides/rpd-talk.qmd")

# The store is passed explicitly to every read, and tar_config_set() is never
# called: that function WRITES a _targets.yaml into the working directory, which is
# how presentations/_targets.yaml came to point at a Windows OneDrive path that has
# not existed for a year.
.store <- here::here("_targets")
.figdir <- "_figuras"
dir.create(.figdir, showWarnings = FALSE, recursive = TRUE)

# Figures are built in a subprocess that dies afterwards ----------------------------
# A ggplot built on rdrobust::rdplot() drags the whole rdplot frame along through the
# plot object's environment -- ~600 MB of RSS per figure once deserialised -- and
# neither gc() nor reassigning plot_env releases it, because the layers' quosures
# keep the reference. Half a dozen of them in one knitr session is an OOM kill on a
# 7.6 GB machine. So no plot object is ever created in this session: each is rendered
# to file in a callr child that exits, and only the image is included here.
#
# This is the amafore/index.qmd pattern with three fixes: `.store` is actually
# defined (that version passes an undefined symbol and survives only because its
# cache is already warm), the child attaches ggpubr (ggarrange objects print through
# print.ggarrange, so ggsave silently produces nothing without it), and staleness is
# keyed on the target's own hash rather than on mere existence.
#
# `nrow`/`ncol` are for the RD panel targets in the backup section. Those targets
# store a *list of panels* rather than an arranged grid, precisely so the deck and the
# working paper can lay the same estimates out differently -- 2x3 in Libertinus Sans
# here, 3x2 in TeX Gyre Heros in sections/. When they are supplied, the child sources
# R/ and calls arrange_rd_panels() with the slide theme.
#
# Every figure is drawn at the size it is displayed at. There used to be a `show`
# argument that let the panel figures be drawn on a big canvas and scaled down, because
# their titles were sized for a page and got clipped at slide size; now that
# `base_size` sets the panel type directly there is nothing to compensate for, and
# rescaling only thinned the strokes.
fig_slides <- function(target, width = 5.5, height = 2.6, res = 300,
                       nrow = NULL, ncol = NULL, base_size = 7) {
  img <- file.path(.figdir, paste0(target, ".png"))
  stamp <- file.path(.figdir, paste0(target, ".hash"))

  # The stamp carries the requested geometry as well as the target's data hash.
  # Keying on the hash alone silently reuses an image drawn at a different size when
  # only the call site changes, which is exactly how a frame ends up overfull by a
  # constant that no amount of resizing removes. The layout belongs in here for the
  # same reason: changing nrow/ncol changes the image without changing the target.
  meta <- targets::tar_meta(names = tidyselect::all_of(target), fields = "data",
                            store = .store)
  # length(meta$data), not nrow(meta): `nrow` is a parameter of this function now, and
  # relying on R skipping a non-function binding in call position is not worth it.
  hash <- paste(
    if (length(meta$data)) meta$data[[1]] else NA_character_,
    width, height, res, nrow %||% "-", ncol %||% "-", base_size
  )
  stale <- !file.exists(img) || !file.exists(stamp) ||
    !identical(readLines(stamp, warn = FALSE), hash)

  if (stale) {
    callr::r(
      function(target, store, dest, width, height, res, nrow, ncol, base_size, root) {
        library(ggplot2)
        library(ggpubr) # ggarrange objects need this to print
        # Libertinus lives only in the texmf tree, so fontconfig cannot see it and
        # cairo devices cannot use it. Registering the OTFs with systemfonts and
        # drawing through ragg is what makes the figures share the slides' typeface.
        kp <- function(f) system2("kpsewhich", f, stdout = TRUE)
        try(systemfonts::register_font(
          "Libertinus Sans",
          plain = kp("LibertinusSans-Regular.otf"),
          bold = kp("LibertinusSans-Bold.otf"),
          italic = kp("LibertinusSans-Italic.otf")
        ), silent = TRUE)
        p <- targets::tar_read_raw(target, store = store)
        if (!is.null(nrow)) {
          # A panel-list target. These three files are pure definitions -- no
          # library() calls, no top-level code needing dplyr -- so sourcing them
          # with only ggplot2 and ggpubr attached is safe. tar_source() is not used
          # because it would pull in the whole R/ tree for two functions.
          for (f in c("rd_functions.R", "results.R", "slides_en.R")) {
            source(file.path(root, "R", f))
          }
          p <- arrange_rd_panels(p, nrow = nrow, ncol = ncol,
                                 theme = theme_slides(base_size = base_size))
        }
        ggplot2::ggsave(dest, p, width = width, height = height,
                        device = ragg::agg_png, res = res, units = "in")
      },
      args = list(target, .store, normalizePath(img, mustWork = FALSE),
                  width, height, res, nrow, ncol, base_size, here::here()),
      libpath = .libPaths()
    )
    writeLines(hash, stamp)
  }
  # Emit the \includegraphics directly rather than going through
  # knitr::include_graphics(). Quarto wraps that in \pandocbounded{} with no width
  # argument, which ignores the dpi entirely and instead rescales the image to
  # \linewidth -- so an intentionally oversized canvas came back 0.27in too tall and
  # overflowed the frame. Stating the width here makes the displayed size exact.
  knitr::asis_output(sprintf(
    "\\noindent\\makebox[\\textwidth][c]{\\includegraphics[width=%.3fin]{%s}}",
    width, img
  ))
}

# Small numeric targets, read once -------------------------------------------------
# None of these loads rd_data; the largest is a few kilobytes. Every number on every
# slide comes from one of them, so the deck cannot drift from the paper.
tr <- function(n) targets::tar_read_raw(n, store = .store)

res_main <- tr("resultados_principales_esp")
het      <- tr("heterogeneidad_esp")
pe       <- tr("primera_etapa_esp")
tu_path  <- tr("take_up_path_data_esp")
muestra  <- tr("muestra_stats_esp")
itt      <- tr("itt_table_esp")
balance  <- tr("balance_table_esp")
bw       <- tr("bw_sensitivity_esp")
shares   <- tr("retiro_saldo_shares_esp")
cost     <- tr("pension_cost_en")
costo_ag <- tr("costo_agregado_esp")
resumen  <- tr("tabla_resumen_esp")
n_retiros <- tr("rpd_usage")
n_usuarios <- tr("rpd_users")

# Accessors, as in iab/abstract.qmd. `het` is keyed on the Spanish subgroup labels of
# the AMAFORE companion paper; only the numeric columns are used here.
rp    <- function(n, col = "coef") res_main[[col]][res_main$name == n]
itv   <- function(n, col = "coef") itt[[col]][itt$name == n]
hetv  <- function(g, s, col)       het[[col]][het$grupo == g & het$subgrupo == s]
difp  <- function(g)               unique(het$dif_pv[het$grupo == g])
fs_pp <- function(m)               tu_path$point[tu_path$months == m] * 100
cst   <- function(col, r = 0.04, c = 0.065) cost[[col]][cost$rate == r & cost$c_rate == c]
agg   <- function(col, r = 0.04, c = 0.065) {
  costo_ag[[col]][costo_ag$tasa == r & costo_ag$c_rate == c]
}

f1 <- function(x) formatC(x, format = "f", digits = 1)
f2 <- function(x) formatC(x, format = "f", digits = 2)
f3 <- function(x) formatC(x, format = "f", digits = 3)
fm <- function(x) formatC(round(x), format = "d", big.mark = ",")
bn <- function(x) formatC(x / 1e9, format = "f", digits = 0, big.mark = ",")
# Tables below are built with escape = FALSE so that $p$ renders as maths rather
# than as three literal characters. That makes a bare % a LaTeX comment, so every
# percentage that reaches a cell has to go through this.
pct <- function(x, d = 1) paste0(formatC(x * 100, format = "f", digits = d), "\\%")

# Slide-local note strings. R/results.R's rdd_footnote_beamer and
# fuzzy_rd_footnote_beamer hard-code "Equation 1" and "Equation 3", which do not
# match this deck's numbering, and editing them would touch the working paper.
note_fuzzy <- paste(
  "Fuzzy RD estimates of the effect of RPD take-up on compliers. Local linear,",
  "triangular kernel, MSE-optimal bandwidth; robust bias-corrected inference.",
  "Currencies in 2024 Mexican pesos."
)
```

## The prediction this paper tests

When a benefit is funded from the worker's own savings, the behavioural response
should be muted: no risk pooling, no fiscal outlay, the worker internalises the cost.
That is what individual accounts were designed to deliver [@feldstein2007], and what
Chile appears to show — its duration response sits in the solidarity-funded tranche,
not in the worker's own account [@hartley2011].

Mexico's *Retiro Parcial por Desempleo* is a clean test: displaced formal workers
withdraw part of their own pension savings, and nothing else changes.

::: {.block}
### The answer
Take-up causes **`r f1(rp("duration_36"))` additional weeks** out of formal employment
over three years, with **no** offsetting gain in wages, job stability or earnings.
:::

Whether that is moral hazard or liquidity decides its welfare content [@chetty2008].

# Setting

## Mexico has no unemployment insurance

Severance is the *de jure* income substitute: 90 days of wages, owed by the employer.
It is *de facto* close to unenforceable [@sadka2024]:

- **Half** of dismissed workers report never receiving it, despite being entitled to
- Only **15%** of dismissed workers whose employers fail to pay severande file a claim
  - A majority of the remaining 85% views the process as either "useless" or "too expensive"
- **30%** of lawsuits are unresolved four years later
- Among workers who *won*, only **20%** collected the full award, while 54% of them failed to collect anything

So the RPD is not one option among several. For most displaced formal workers it is
the only income substitute that actually arrives.

## What the RPD is

::: {.columns}
::: {.column width="52%"}
Withdraw part of your pension savings after **46 days** without a registered formal
job.

**Scheme A** — 3 years since entry *and* 2 years of contributions. Pays the lesser of
10 months of minimum wage or one month of own wage.

**Scheme B** — 5 years since entry, no contribution test. Pays the lesser of 3 monthly
wages or 11.5% of the balance.

Usable once every 5 years. Under Art. 198 LSS the withdrawal also **discounts
contribution weeks in the same proportion** as the balance.
:::
::: {.column width="48%"}
```{r}
#| label: fig-elig
fig_slides("eligibility_plot_slides", width = 2.5, height = 2.2)
```
:::
:::

## The program is large, and its use is crisis-driven

```{r}
#| label: fig-usage
fig_slides("rpd_usage_plot_slides", width = 5.2, height = 1.82)
```

`r fm(n_retiros)` withdrawals by `r fm(n_usuarios)` workers since 1997, totalling
\$`r bn(agg("monto_2024"))` bn. The steps are 2009 and 2020: rate of adoption increases when the labor
market deteriorates.

## Data and sample

::: {.columns}
::: {.column width="52%"}
Two confidential IMSS sources linked by social-security number: the matched
employer–employee panel (14.8 M workers, 406.8 M contracts) and the universe of RPD
withdrawal records.

Analysis sample: **`r fm(muestra$n_muestra)`** workers with exactly **one** non-employment spell of
46+ days *during their third year* in formal work.

The withdrawal is small at the moment of use — \$`r fm(muestra$w_retiro)`, about
`r f2(muestra$w_retiro / muestra$wbar)` months of the prior wage, a median of
`r f1(shares$p50_total[1] * 100)`% of the balance.

**Scope condition:** these data observe *formal* employment and only formal employment.
:::
::: {.column width="48%"}
```{r}
#| label: fig-sample
fig_slides("sample_selection_plot_slides", width = 2.5, height = 2.2)
```
:::
:::

# Design

## Fuzzy regression discontinuity

Eligibility jumps at two years of contributions; take-up does not jump all the way.
So eligibility instruments for take-up.

$$
T_{i} = \alpha + \beta D_i + f(C_i - 2) + u_i, \qquad
Y_i = \gamma + \tau \hat T_i + f(C_i - 2) + v_i
$$

for $|C_i - 2| \le h$, where $D_i = \mathbb{1}\{C_i \ge 2\}$.

$\tau$ is the local average treatment effect on **compliers** — workers who withdraw
because the rule let them, and would not have otherwise.

Local linear, triangular kernel, MSE-optimal bandwidth [@calonico2020], robust
bias-corrected inference [@calonico2014].

## First stage

::: {.columns}
::: {.column width="55%"}
```{r}
#| label: fig-fs
fig_slides("rd_take_up_plot_slides", width = 2.8, height = 2.2)
```
:::
::: {.column width="45%"}
Crossing the threshold raises take-up within 12 months by
**`r f1(pe$coef * 100)` pp** (s.e. `r f1(pe$se * 100)`).

$F = `r round(pe$f)`$, on `r fm(pe$n_efectivo)` observations inside a bandwidth of
±`r round(pe$h)` days.

Strong instrument, small effect — which is what a self-financed, partial withdrawal
from your own account should look like.
:::
:::

## Take-up over the horizon

```{r}
#| label: fig-fspath
fig_slides("take_up_path_slides", width = 5.2, height = 1.6)
```

`r f1(fs_pp(2))` pp at two months, `r f1(fs_pp(12))` at twelve, `r f1(fs_pp(24))` by
month 24. The flattening after month 12 is **not** non-compliance: ineligible workers
reach Scheme B's five-year threshold and become eligible on their own.

## Design validity

::: {.columns}
::: {.column width="52%"}
```{r}
#| label: fig-density
fig_slides("density_plot_slides", width = 2.5, height = 2.1)
```
:::
::: {.column width="48%"}
**No manipulation** [@cattaneo2020]. Landing on the eligible side means postponing
your own dismissal by days — not a margin workers control.

**Balance.** Of `r nrow(balance)` covariates, `r sum(balance$pv < 0.05)` reject at 5%
individually; under Holm **none** survives. The largest jump is age — about 1.5 months
against a mean of `r f1(muestra$edad_media)` years.

**Bandwidth.** Positive across ×0.5–×2 of the optimal width.
:::
:::

## How to read what follows

The first stage *is* the complier share: about **4 in 100** workers near the threshold.

The LATE describes *them* — not the average worker, and not the average program user.
Two consequences:

- Effects look large against population means, because they are concentrated in a
  small and specific group.
- The **reduced form is correspondingly small**: the ITT on 36-month duration is
  `r f2(itv("duration_36"))` weeks ($p$ = `r f2(itv("duration_36", "pv"))`).

The reduced form is not in tension with the LATE. It is the same evidence expressed
per capita rather than per user.

# Results

## Take-up lengthens time out of formal work

```{r}
#| label: fig-fuzzypath
fig_slides("fuzzy_path_slides", width = 5.2, height = 2.25)
```

## Take-up lengthens time out of formal work

```{r}
#| label: tbl-duration
dur <- tibble::tibble(
  Outcome = c("Still out of formal work at 3 months (pp)",
              "Weeks out of formal work over 36 months"),
  Effect = c(f1(rp("survival_3") * 100), f1(rp("duration_36"))),
  SE = paste0("(", c(f1(rp("survival_3", "se") * 100), f1(rp("duration_36", "se"))), ")"),
  p = c(f3(rp("survival_3", "pv")), f3(rp("duration_36", "pv"))),
  Base = c(pct(rp("survival_3", "media")), f1(rp("duration_36", "media")))
)

kbl(dur, format = "latex", booktabs = TRUE, escape = FALSE, longtable = FALSE,
      linesep = "",
    col.names = c("Outcome", "Effect", "Std. err.", "$p$", "Ineligible mean"),
    align = c("l", "r", "r", "r", "r")) |>
  kable_styling(font_size = 8, latex_options = "hold_position") |>
  column_spec(1, width = "5cm") |>
  column_spec(2:5, width = "1.7cm")
```

Both significant at 5%. The 36-month effect is `r round(rp("duration_36") / rp("duration_36", "media") * 100)`%
of the `r f1(rp("duration_36", "media"))` weeks an ineligible worker spends out of
formal work — a reference for scale, not a rate: the two are not additive, because
compliers' spells are plausibly longer to begin with.

Consistent with the UI literature, and with lump-sum severance in Brazil
[@britto2022].

## Nothing compensates the extra weeks

```{r}
#| label: tbl-nogain
gains <- tibble::tibble(
  Outcome = c("Monthly earnings, next job",
              "Total earnings, next job",
              "Next job duration (weeks)",
              "Cumulative earnings, 3 years",
              "Months worked, 3 years"),
  key = c("next_job_av_earnings", "next_job_cum_earnings", "next_job_duration",
          "earnings_total", "months_worked_total")
) |>
  mutate(
    Effect = map_chr(key, ~ fm(rp(.x))),
    SE = map_chr(key, ~ paste0("(", fm(rp(.x, "se")), ")")),
    p = map_chr(key, ~ f3(rp(.x, "pv"))),
    Base = map_chr(key, ~ fm(rp(.x, "media")))
  ) |>
  select(-key)

kbl(gains, format = "latex", booktabs = TRUE, escape = FALSE, longtable = FALSE,
      linesep = "",
    col.names = c("Outcome", "Effect", "Std. err.", "$p$", "Ineligible mean"),
    align = c("l", "r", "r", "r", "r")) |>
  kable_styling(font_size = 8, latex_options = "hold_position") |>
  column_spec(1, width = "4.5cm") |>
  column_spec(2:5, width = "1.9cm") |>
  footnote(general = note_fuzzy, general_title = "Note: ",
           threeparttable = TRUE, escape = FALSE, footnote_as_chunk = TRUE)
```

The next-job wage estimate is sizeable but misses 5% ($p$ = `r f3(rp("next_job_av_earnings", "pv"))`);
everything else is zero. Against @acemoglu2000; with @britto2022, @gonzález-rozada2023.

## Who pays — and which contrasts survive a test

```{r}
#| label: fig-het
fig_slides("het_diff_plot_slides", width = 5.2, height = 2.25)
```

## Who pays — and which contrasts survive a test

Each split is estimated on **disjoint** subsamples, each with its own bandwidth. A
large coefficient in one half and a small one in the other does *not* establish that
the halves differ. So test the difference directly:

- **Prior earnings.** Below median `r round(hetv("Ingreso previo", "Por debajo de la mediana", "dur_coef"))` weeks
  vs. above median `r f1(hetv("Ingreso previo", "Por arriba de la mediana", "dur_coef"))`.
  Difference $p$ = `r f3(difp("Ingreso previo"))` — **distinguishable**.
- **Sex.** Men `r round(hetv("Sexo", "Hombres", "dur_coef"))` weeks vs. women
  `r f1(hetv("Sexo", "Mujeres", "dur_coef"))`. Difference $p$ = `r f3(difp("Sexo"))` — **distinguishable**.
- **Age** ($p$ = `r f3(difp("Edad"))`) and **COVID-19 exposure** ($p$ = `r f2(difp("Exposición a COVID-19"))`)
  point the same way but are **not** distinguishable. Suggestive, not findings.

No subgroup shows a detectable three-year earnings gain. The *only* reemployment-wage
gain significant at 5% accrues to workers already **above** the median
(\$`r fm(hetv("Ingreso previo", "Por arriba de la mediana", "wage_coef"))`,
$p$ = `r f3(hetv("Ingreso previo", "Por arriba de la mediana", "wage_pv"))`).

## Longer search, or exit into informality?

These data cannot tell the two apart. A worker who withdraws and then works informally
or on her own account looks exactly like one still searching.

There is reason to think the second margin is real: @britto2022 finds 57% of the
displaced formal employment offset informally; @gerard2021 shows the efficiency cost
is lower where formal reentry is unlikely; and for Mexico, @carreñogodínez2025 finds
RPD eligibility raises transitions into **self-employment**.

::: {.block}
### The retirement conclusion is invariant
If much of the `r f1(rp("duration_36"))` weeks is informal work, the *consumption*
cost is smaller than it looks. The *pension* cost is identical either way: neither
informal work nor self-employment generates contributions or accrues weeks.
:::

# What it costs at retirement

## Two channels, one compounding factor

::: {.columns}
::: {.column width="46%"}
**Channel 1** — the withdrawal stops compounding:
$W(1+r)^n$

**Channel 2** — contributions never made during the induced non-employment:
$\bar{w} \cdot c \cdot \Delta \cdot (1+r)^n$

Both compound at the same rate, so channel 2's *share* is invariant to $r$ — no need
to take a stand on returns.
:::
::: {.column width="54%"}
```{r}
#| label: fig-channels
fig_slides("fig_pension_channels_en", width = 2.8, height = 1.6)
```
:::
:::

At the sample's own values (\$`r fm(cst("withdrawal"))` withdrawn, \$`r fm(cst("wage"))`
monthly wage, `r f1(cst("weeks"))` extra weeks, `r f1(0.065 * 100)`% contribution rate,
40 years): the loss is **`r f1(min(cost$multiple[cost$c_rate == 0.065]))`–`r f1(max(cost$multiple[cost$c_rate == 0.065]))`×
the amount withdrawn**, `r round(cst("share2") * 100)`% of it from the unemployment.

## The burden is regressive in the ratio

```{r}
#| label: fig-quintiles
#| layout-ncol: 2
fig_slides("fig_retiro_saldo_en", width = 2.5, height = 1.78)
fig_slides("fig_retiro_semanas_en", width = 2.5, height = 1.78)
```

Low-wage users withdraw a **larger share** of a smaller balance:
`r f1(shares$prop_25[1] * 100)`% of Q1 withdrawals take a quarter or more of the account,
against `r f1(shares$prop_25[5] * 100)`% in Q5.

# Takeaways

## Takeaways

1. **The individual-account prediction does not survive.** A benefit funded entirely
   from the worker's own savings still produces a large duration response —
   `r f1(rp("duration_36"))` weeks — with no matching gain.

2. **The pattern looks like liquidity, not classical moral hazard.** A large response,
   no compensating improvement, and a cost concentrated among below-median earners and
   among men — the two contrasts that survive a formal test.

3. **The cost is paid in the pension, and most of it is invisible.**
   `r f1(cst("multiple"))`× the withdrawal at a 4% real return, of which
   `r round(cst("share2") * 100)`% comes from the induced unemployment rather than the
   withdrawal itself.

4. **The cost moves with the cycle; the program's parameters do not.** How much
   reemployment the liquidity displaces depends on how much reemployment is available
   to displace [@chetty2008].

## Limitations

- **External validity.** The LATE describes compliers displaced in their third year of
  formal work. Not the average user, not the average worker.
- **Informality is unobserved.** See the discussion above; it bounds the consumption
  interpretation but not the pension one.
- **Precision.** The second-stage confidence intervals are wide
  (36-month duration: `r f1(rp("duration_36") - 1.96 * rp("duration_36", "se"))`–`r f1(rp("duration_36") + 1.96 * rp("duration_36", "se"))` weeks).
- **Control-group contamination.** Ineligible workers become eligible under Scheme B
  within the three-year window, so long-horizon estimates are a lower bound.



# Backup

```{r}
#| label: summary-prep
#| include: false

# tabla_resumen_esp is the AMAFORE companion paper's summary target: Spanish labels,
# raw numerics. Translate the labels here and pick a formatter per row, since the
# same column mixes proportions, calendar years, weeks and pesos.
resumen_en <- tibble::tribble(
  ~etiqueta,                                      ~Variable,                              ~kind,
  "Mujer (proporción)",                           "Female",                               "prop",
  "Fecha de nacimiento (año)",                    "Year of birth",                        "year",
  "Inicio de vida laboral formal (año)",          "Year of entry to formal work",         "year",
  "Fecha de separación (año)",                    "Year of displacement",                 "year",
  "Edad a la separación (años)",                  "Age at displacement",                  "num",
  "Días desde la apertura de la cuenta",          "Days since account opened",            "int",
  "Sin CURP (proporción)",                        "No CURP",                              "prop",
  "Duración del empleo previo (semanas)",         "Previous job duration (weeks)",        "num",
  "Salario mensual, empleo previo (MXN 2024)",    "Previous job, monthly wage",           "int",
  "Ingresos totales, empleo previo (MXN 2024)",   "Previous job, total earnings",         "int",
  "Monto retirado (MXN 2024)",                    "Amount withdrawn",                     "int",
  "Saldo RCV al momento del retiro (MXN 2024)",   "RCV balance at withdrawal",            "int",
  "Semanas cotizadas al momento del retiro",      "Contribution weeks at withdrawal",     "num",
  "Días entre la separación y el retiro",         "Days from displacement to withdrawal", "int",
  "Días de salario retirados",                    "Days of wages withdrawn",              "int",
  "Sigue desempleado a 3 meses (proporción)",     "Still out of formal work, 3 months",   "prop",
  "Sigue desempleado a 36 meses (proporción)",    "Still out of formal work, 36 months",  "prop",
  "Semanas de desempleo, censuradas a 3 meses",   "Weeks out of formal work, 3 months",   "num",
  "Semanas de desempleo, censuradas a 36 meses",  "Weeks out of formal work, 36 months",  "num",
  "Duración del siguiente empleo (semanas)",      "Next job duration (weeks)",            "num",
  "Salario mensual, siguiente empleo (MXN 2024)", "Next job, monthly wage",               "int",
  "Ingresos totales, siguiente empleo (MXN 2024)","Next job, total earnings",             "int",
  "Ingresos totales a 3 años (MXN 2024)",         "Cumulative earnings, 3 years",         "int",
  "Salario mensual promedio a 3 años (MXN 2024)", "Average monthly wage, 3 years",        "int",
  "Meses trabajados a 3 años",                    "Months worked, 3 years",               "int"
)

fmt_kind <- function(x, kind) {
  out <- character(length(x))
  out[kind == "prop"] <- f3(x[kind == "prop"])
  out[kind == "year"] <- formatC(x[kind == "year"], format = "f", digits = 1)
  out[kind == "num"]  <- f1(x[kind == "num"])
  out[kind == "int"]  <- fm(x[kind == "int"])
  out
}

summary_slide <- function(groups) {
  resumen |>
    filter(grupo %in% groups) |>
    left_join(resumen_en, by = "etiqueta") |>
    mutate(
      Mean = fmt_kind(media, kind),
      Median = fmt_kind(mediana, kind),
      SD = fmt_kind(de, kind),
      N = fm(n)
    ) |>
    select(Variable, Mean, Median, SD, N) |>
    kbl(format = "latex", booktabs = TRUE, escape = FALSE, longtable = FALSE,
      linesep = "",
        col.names = c("Variable", "Mean", "Median", "Std. dev.", "N"),
        align = c("l", "r", "r", "r", "r")) |>
    kable_styling(font_size = 7, latex_options = "hold_position") |>
    column_spec(1, width = "5.0cm") |>
    column_spec(2:5, width = "1.5cm")
}
```

## Summary statistics: covariates and previous job

```{r}
#| label: tbl-summary
summary_slide(c("Covariables", "Empleo previo"))
```

## Summary statistics: the withdrawal and job search

```{r}
#| label: tbl-summary2
summary_slide(c("Retiro (RPD)", "Búsqueda de empleo"))
```

Withdrawal variables are defined on the subsample that used the program.

## Summary statistics: next job and medium term

```{r}
#| label: tbl-summary3
summary_slide(c("Siguiente empleo", "Mediano plazo"))
```

Monetary values in 2024 Mexican pesos.

## Covariate balance, with the Holm correction

```{r}
#| label: tbl-balance
balance |>
  mutate(
    holm = p.adjust(pv, method = "holm"),
    Covariate = c("Female", "Date of birth", "Entry to formal work",
                  "Date of displacement", "Age", "Days since account opened",
                  "No CURP", "Previous job duration (wks)",
                  "Previous job total earnings", "Previous job monthly wage"),
    Effect = fm(coef), SE = paste0("(", fm(se), ")"),
    p = f3(pv), Holm = f3(holm)
  ) |>
  select(Covariate, Effect, SE, p, Holm) |>
  kbl(format = "latex", booktabs = TRUE, escape = FALSE, longtable = FALSE,
      linesep = "",
      col.names = c("Covariate", "Effect", "Std. err.", "$p$", "Holm $p$"),
      align = c("l", "r", "r", "r", "r")) |>
  kable_styling(font_size = 7, latex_options = "hold_position") |>
  column_spec(1, width = "4.0cm") |>
  column_spec(2:5, width = "1.9cm") |>
  footnote(general = paste("Sharp RD estimates of the eligibility effect on each",
                           "covariate. Holm-adjusted p-values across the 10 tests."),
           general_title = "Note: ", threeparttable = TRUE, escape = FALSE,
           footnote_as_chunk = TRUE)
```

## Reduced-form (ITT) estimates

```{r}
#| label: tbl-itt
itt |>
  mutate(
    Outcome = c("RPD take-up at 12 months (first stage)",
                "Still out of formal work at 3 months",
                "Weeks out of formal work, 36 months",
                "Monthly earnings, next job",
                "Total earnings, next job",
                "Next job duration (weeks)",
                "Cumulative earnings, 3 years",
                "Months worked, 3 years"),
    Effect = ifelse(abs(coef) < 1, f3(coef), fm(coef)),
    SE = paste0("(", ifelse(abs(se) < 1, f3(se), fm(se)), ")"),
    p = f3(pv), N = fm(n_efectivo)
  ) |>
  select(Outcome, Effect, SE, p, N) |>
  kbl(format = "latex", booktabs = TRUE, escape = FALSE, longtable = FALSE,
      linesep = "",
      col.names = c("Outcome", "Effect", "Std. err.", "$p$", "Eff. N"),
      align = c("l", "r", "r", "r", "r")) |>
  kable_styling(font_size = 7, latex_options = "hold_position") |>
  column_spec(1, width = "4.6cm") |>
  column_spec(2:5, width = "1.9cm")
```

No reduced-form effect is significant at 5% — what a `r f1(pe$coef * 100)` pp first
stage implies. Reduced form ÷ first stage ≠ the LATE: each estimate picks its own
bandwidth.

## Sensitivity to the bandwidth

```{r}
#| label: fig-bw
fig_slides("bw_plot_slides", width = 4.9, height = 1.82)
```

Positive throughout; significance concentrates near the optimal bandwidth, and the
estimate attenuates as the window widens.

## Discontinuities in the two main outcomes

```{r}
#| label: fig-rdoutcome
fig_slides("rd_outcome_plot_slides", width = 5.2, height = 2.25)
```

## Take-up at every horizon

```{r}
#| label: fig-takeup-panels
fig_slides("take_up_plots", width = 5.2, height = 2.6, nrow = 2, ncol = 3)
```

## Covariate continuity

```{r}
#| label: fig-cov
fig_slides("covariates_plots", width = 5.2, height = 2.6, nrow = 2, ncol = 3)
```

## Previous-job continuity

```{r}
#| label: fig-prevjob
fig_slides("prev_job_plots", width = 4.4, height = 2.6, nrow = 2, ncol = 2)
```

## Survival in non-employment, by horizon

```{r}
#| label: fig-surv
fig_slides("survival_plots", width = 5.2, height = 2.6, nrow = 2, ncol = 3)
```

## Duration of non-employment, by horizon

```{r}
#| label: fig-dur
fig_slides("duration_plots", width = 5.2, height = 2.6, nrow = 2, ncol = 3)
```

## Heterogeneity: the underlying estimates

```{r}
#| label: tbl-het
het |>
  mutate(
    Split = unname(c("Ingreso previo" = "Prior earnings", "Edad" = "Age",
                     "Sexo" = "Sex", "Exposición a COVID-19" = "COVID-19")[grupo]),
    Subgroup = unname(c(
      "Por debajo de la mediana" = "Below median",
      "Por arriba de la mediana" = "Above median",
      "Hombres" = "Men", "Mujeres" = "Women",
      "Separación antes de marzo 2019" = "Before Mar 2019",
      "Separación después de marzo 2020" = "After Mar 2020")[subgrupo]),
    Duration = paste0(f1(dur_coef), " (", f1(dur_se), ")"),
    `Dur. p` = f3(dur_pv),
    Earnings = paste0(fm(ing_coef), " (", fm(ing_se), ")"),
    Wage = paste0(fm(wage_coef), " (", fm(wage_se), ")"),
    `Diff. p` = f3(dif_pv)
  ) |>
  select(Split, Subgroup, Duration, `Dur. p`, Earnings, Wage, `Diff. p`) |>
  kbl(format = "latex", booktabs = TRUE, escape = FALSE, longtable = FALSE,
      linesep = "",
      align = c("l", "l", "r", "r", "r", "r", "r")) |>
  kable_styling(font_size = 6.5, latex_options = "hold_position") |>
  footnote(general = paste(
    "Duration is weeks out of formal work over 36 months; Earnings is cumulative",
    "three-year earnings; Wage is monthly earnings in the next job. Standard errors",
    "in parentheses. Diff. p tests the difference between the two halves of each",
    "split on the duration outcome; subsamples are disjoint, so the variance of the",
    "difference is the sum of the variances."),
    general_title = "Note: ", threeparttable = TRUE, escape = FALSE,
    footnote_as_chunk = TRUE)
```

## Pension cost: the full grid

```{r}
#| label: tbl-pension
cost |>
  mutate(
    Rate = pct(rate, 0),
    `Contrib. rate` = pct(c_rate, 1),
    `Channel 1` = fm(channel1), `Channel 2` = fm(channel2),
    Total = fm(total), Multiple = paste0(f1(multiple), "x"),
    `Ch. 2 share` = pct(share2, 1)
  ) |>
  select(Rate, `Contrib. rate`, `Channel 1`, `Channel 2`, Total, Multiple,
         `Ch. 2 share`) |>
  kbl(format = "latex", booktabs = TRUE, escape = FALSE, longtable = FALSE,
      linesep = "",
      align = c("r", "r", "r", "r", "r", "r", "r")) |>
  kable_styling(font_size = 7, latex_options = "hold_position") |>
  footnote(general = paste(
    "Illustrative, not actuarial: constant real return, no fees, no partial recovery",
    "of contributions after reemployment. It combines a complier LATE (the extra",
    "weeks) with sample means (the withdrawal and the prior wage), so it describes a",
    "worker of average characteristics induced by the eligibility rule. The 6.5%",
    "contribution rate is the one in force over the estimation window; 15% is the",
    "2020 reform ceiling."),
    general_title = "Note: ", threeparttable = TRUE, escape = FALSE,
    footnote_as_chunk = TRUE)
```

## Aggregate cost of the program to date

Applying the same two channels to the whole withdrawal registry, at a
`r f1(0.04 * 100)`% real return and the `r f1(0.065 * 100)`% contribution rate:

- \$`r bn(agg("monto_2024"))` bn withdrawn (2024 pesos), across `r fm(agg("n_retiros"))` withdrawals
- \$`r bn(agg("canal1"))` bn once capitalised — channel 1
- \$`r bn(agg("canal2"))` bn of contributions never made — channel 2
- **\$`r bn(agg("total"))` bn total**, a markup of `r f2(agg("markup"))`× on the capitalised withdrawals alone

Note the channel-2 share here is `r round(agg("canal2") / agg("total") * 100)`%, not
the `r round(cst("share2") * 100)`% of the individual calculation: the registry uses
its own wage distribution rather than the estimation sample's mean, and the estimation
sample sits at the two-year threshold where withdrawals are small relative to wages.


## References {.allowframebreaks}

::: {#refs}
:::
