---
title: "Fixed-Bandwidth Robustness of the RPD RD Estimates"
subtitle: "Holding the estimation window constant across horizons"
format:
  html:
    toc: true
    embed-resources: true
fig-width: 9
fig-height: 5
execute:
  cache: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| output: false

# renv does not auto-activate when Quarto starts R inside notebooks/ —
# walk up to the project root and activate it if needed.
if (!requireNamespace("pacman", quietly = TRUE) ||
    !requireNamespace("rdrobust", quietly = TRUE)) {
  root <- normalizePath(getwd())
  while (!file.exists(file.path(root, "renv", "activate.R")) &&
         dirname(root) != root) {
    root <- dirname(root)
  }
  Sys.setenv(RENV_PROJECT = root)
  source(file.path(root, "renv", "activate.R"))
}

pacman::p_load(tidyverse, here, rdrobust, targets, ggpubr, gt)

here::i_am("notebooks/fixed_bw.qmd")

# Read-only reuse of the pipeline's estimators (my_rd, my_fuzzy_rd).
source(here("R/rd_functions.R"))

theme_set(theme_minimal())

# Explicit store: notebooks/_targets.yaml is stale (old OneDrive path).
store <- here("_targets")

family_levels <- c("take_up", "survival", "duration")
family_labels <- c(
  take_up  = "Take-up (first stage)",
  survival = "Survival",
  duration = "Unemployment duration"
)

# Robust-row stats from an rdrobust fit wrapped as list(output = ...).
extract_stats <- function(fit) {
  out <- fit$output
  tibble(
    h     = out$bws[1],
    b     = out$bws[2],
    n_eff = sum(out$N_h),
    coef  = out$coef[3],
    se    = out$se[3],
    pv    = out$pv[3],
    ci_l  = out$ci[3],
    ci_u  = out$ci[6]
  )
}

# Fit at a manual bandwidth, preserving the anchor's rho = b0 / h0.
# Returns only the stats row so the rdrobust object is freed immediately.
fit_manual <- function(d, h, rho, fuzzy = FALSE) {
  fit <- if (fuzzy) {
    my_fuzzy_rd(d, h = h, b = h * rho)
  } else {
    my_rd(d, h = h, b = h * rho)
  }
  extract_stats(fit)
}
```

```{r}
#| label: load-data
#| output: false

# Pipe tar_read() straight into filter() so the full ~3.7 GB rd_data tibble
# is never bound to a name; keep only the three outcome families.
rd_use <- tar_read(rd_data, store = store) |>
  filter(str_detect(name, "^(take_up|survival|duration)_\\d+$")) |>
  select(name, data) |>
  mutate(
    months = str_extract(name, "\\d+") |> as.numeric(),
    family = str_remove(name, "_\\d+$") |> factor(levels = family_levels)
  )
invisible(gc())
```

## Motivation

The main results of the paper let `rdrobust` choose the MSE-optimal bandwidth
(Calonico, Cattaneo, and Titiunik 2014) separately for every regression. Each
estimate is individually valid, but within an outcome family measured at many
horizons (take-up, survival, unemployment duration at 2–36 months after
displacement), a bandwidth that wanders across horizons means each point of the
path is estimated on a different local sample. This notebook re-estimates the
first stage, survival, and duration paths **holding the bandwidth fixed within
each family**, and shows how sensitive the anchor estimates are to the
bandwidth choice.

All regressions inherit the paper's specification: local linear regression,
triangular kernel, HC0 variance, robust bias-corrected inference. When the main
bandwidth $h$ is fixed manually, the bias bandwidth $b$ is scaled to preserve
the anchor regression's ratio $\rho = b_0 / h_0$.

```{r}
#| label: anchors
#| output: false

anchor_data <- list(
  take_up  = rd_use |> filter(name == "take_up_12")  |> pluck("data", 1),
  survival = rd_use |> filter(name == "survival_24") |> pluck("data", 1),
  duration = rd_use |> filter(name == "duration_36") |> pluck("data", 1)
)

anchor_fits <- tribble(
  ~family,    ~anchor,       ~estimand,     ~fuzzy,
  "take_up",  "take_up_12",  "First stage", FALSE,
  "survival", "survival_24", "ITT",         FALSE,
  "survival", "survival_24", "LATE",        TRUE,
  "duration", "duration_36", "ITT",         FALSE,
  "duration", "duration_36", "LATE",        TRUE
) |>
  mutate(
    stats = map2(family, fuzzy, \(fam, fz) {
      fit <- if (fz) my_fuzzy_rd(anchor_data[[fam]]) else my_rd(anchor_data[[fam]])
      extract_stats(fit)
    }),
    family = factor(family, levels = family_levels)
  ) |>
  unnest(stats)

# Fixed-h choice: sharp anchor for the first stage; fuzzy (LATE) anchors for
# survival and duration — LATE is the headline estimand. Round h to the
# nearest 5 days; one (h, rho) per family, shared by ITT and LATE.
bw_spec <- anchor_fits |>
  filter(estimand %in% c("First stage", "LATE")) |>
  transmute(
    family,
    h0    = h,
    b0    = b,
    rho   = b0 / h0,
    h_fix = round(h0 / 5) * 5,
    b_fix = h_fix * rho
  )

# Consistency check: the stored pipeline fits should match fresh fits on the
# current rd_data (guards against a stale _targets store).
stored_late_24 <- tar_read(survival_iv, store = store) |>
  filter(name == "survival_24") |>
  pluck("rd", 1, "output", "coef", 3)
fresh_late_24 <- anchor_fits |>
  filter(anchor == "survival_24", estimand == "LATE") |>
  pull(coef)
if (!isTRUE(all.equal(stored_late_24, fresh_late_24))) {
  warning(
    "Stored pipeline fits differ from fresh fits on current rd_data; ",
    "optimal-h overlays may be stale."
  )
}
rm(stored_late_24, fresh_late_24)
```

```{r}
#| label: sensitivity
#| output: false

h_grid <- seq(30, 200, by = 10)

sens <- tribble(
  ~family,    ~estimand,     ~fuzzy,
  "take_up",  "First stage", FALSE,
  "survival", "ITT",         FALSE,
  "survival", "LATE",        TRUE,
  "duration", "ITT",         FALSE,
  "duration", "LATE",        TRUE
) |>
  left_join(bw_spec |> mutate(family = as.character(family)), by = "family") |>
  cross_join(tibble(h_manual = h_grid)) |>
  mutate(
    stats = pmap(
      list(family, h_manual, rho, fuzzy),
      \(fam, h, r, fz) fit_manual(anchor_data[[fam]], h, r, fuzzy = fz),
      .progress = TRUE
    ),
    family = factor(family, levels = family_levels)
  ) |>
  unnest(stats)
```

```{r}
#| label: fixed-paths
#| output: false

with_spec <- rd_use |>
  left_join(bw_spec |> select(family, h_fix, rho), by = "family")

paths_sharp <- with_spec |>
  mutate(
    stats = pmap(list(data, h_fix, rho), fit_manual, .progress = TRUE),
    estimand = if_else(family == "take_up", "First stage", "ITT")
  ) |>
  select(family, name, months, estimand, stats)

paths_late <- with_spec |>
  filter(family != "take_up") |>
  mutate(
    stats = pmap(
      list(data, h_fix, rho),
      \(d, h, r) fit_manual(d, h, r, fuzzy = TRUE),
      .progress = TRUE
    ),
    estimand = "LATE"
  ) |>
  select(family, name, months, estimand, stats)

paths_fixed <- bind_rows(paths_sharp, paths_late) |> unnest(stats)

# Everything downstream is KB-sized; drop the ~2.9 GB of nested data.
rm(rd_use, with_spec, paths_sharp, paths_late, anchor_data)
invisible(gc())
```

```{r}
#| label: stored-paths
#| output: false

# Optimal-bandwidth paths come from the stored pipeline fits — no re-fitting.
extract_stored <- function(tbl) {
  tbl |>
    mutate(stats = map(rd, extract_stats)) |>
    select(name, stats) |>
    unnest(stats) |>
    mutate(
      months = str_extract(name, "\\d+") |> as.numeric(),
      family = str_remove(name, "_\\d+$") |> factor(levels = family_levels)
    )
}

paths_opt <- bind_rows(
  tar_read(take_up, store = store) |>
    extract_stored() |>
    mutate(estimand = "First stage"),
  tar_read(survival, store = store) |>
    extract_stored() |>
    mutate(estimand = "ITT"),
  tar_read(survival_iv, store = store) |>
    extract_stored() |>
    mutate(estimand = "LATE")
)
```

## How much does the optimal bandwidth wander?

@fig-h-by-horizon plots the MSE-optimal bandwidth chosen by `rdrobust` at every
horizon of each outcome family, as used in the paper's path figures. The dashed
line marks the fixed bandwidth adopted for that family in this notebook.

```{r}
#| label: fig-h-by-horizon
#| fig-cap: "CCT-optimal bandwidth by horizon. Dashed line: fixed h adopted for the family."

paths_opt |>
  ggplot(aes(months, h, color = estimand)) +
  geom_line() +
  geom_point(size = 0.8) +
  geom_hline(
    data = bw_spec,
    aes(yintercept = h_fix),
    linetype = "dashed"
  ) +
  facet_wrap(~family, labeller = labeller(family = family_labels)) +
  scale_x_continuous(breaks = seq(3, 36, 3)) +
  labs(
    x = "Months since displacement",
    y = "Optimal bandwidth h (days)",
    color = NULL
  )
```

## Anchor regressions and the fixed-bandwidth choice

One anchor horizon per family — the first stage at 12 months, survival at 24
months, duration at 36 months — is estimated at the CCT-optimal bandwidth. The
fixed bandwidth is the anchor's $h_0$ rounded to the nearest 5 days (sharp
anchor for the first stage, fuzzy anchors for survival and duration), applied
to every horizon in the family for both ITT and LATE.

```{r}
#| label: tbl-anchors
#| tbl-cap: "Anchor regressions at the CCT-optimal bandwidth and the fixed h adopted per family."

anchor_fits |>
  left_join(bw_spec |> select(family, h_fix), by = "family") |>
  mutate(family = family_labels[as.character(family)]) |>
  select(
    Family = family, Anchor = anchor, Estimand = estimand,
    `h (days)` = h, `b (days)` = b, `Effective N` = n_eff,
    Estimate = coef, SE = se, `p-value` = pv, `Fixed h (days)` = h_fix
  ) |>
  gt(groupname_col = "Family") |>
  fmt_number(columns = c(`h (days)`, `b (days)`), decimals = 1) |>
  fmt_number(columns = `Effective N`, decimals = 0, use_seps = TRUE) |>
  fmt_number(columns = c(Estimate, SE, `p-value`), decimals = 3) |>
  tab_source_note(
    "Local linear regression, triangular kernel, HC0 variance. Robust
     bias-corrected estimates and inference (Calonico, Cattaneo, and
     Titiunik 2014). Effective N counts observations within the bandwidth."
  )
```

## Bandwidth sensitivity at the anchor horizons

Each panel re-estimates the anchor regression over a grid of fixed bandwidths
$h \in \{30, 40, \ldots, 200\}$ days, scaling $b$ to preserve the anchor's
$\rho$. Dashed vertical line: CCT-optimal $h_0$; solid line: adopted fixed $h$.
At large $h$ the scaled $b$ exceeds the running variable's $\pm 364$-day
support, so the bias regression effectively uses the full sample — expected and
harmless.

```{r}
#| label: fig-sens-plots
#| output: false

plot_sens <- function(fam) {
  df <- sens |> filter(family == fam)
  vl <- anchor_fits |> filter(family == fam) |> select(estimand, h0 = h)
  fx <- bw_spec |> filter(family == fam)

  ggplot(df, aes(h_manual, coef)) +
    geom_hline(yintercept = 0, color = "gray60") +
    geom_vline(
      data = vl, aes(xintercept = h0),
      linetype = "dashed", color = "gray40"
    ) +
    geom_vline(xintercept = fx$h_fix, color = "steelblue") +
    geom_pointrange(aes(ymin = ci_l, ymax = ci_u), size = 0.3) +
    facet_wrap(~estimand, scales = "free_y") +
    labs(x = "Bandwidth h (days)", y = "RD estimate (robust)")
}
```

```{r}
#| label: fig-sens-takeup
#| fig-cap: "First stage at 12 months (take_up_12): estimate vs. fixed bandwidth."

plot_sens("take_up")
```

```{r}
#| label: fig-sens-survival
#| fig-cap: "Survival at 24 months (survival_24): estimate vs. fixed bandwidth, ITT and LATE."

plot_sens("survival")
```

```{r}
#| label: fig-sens-duration
#| fig-cap: "Duration at 36 months (duration_36): estimate vs. fixed bandwidth, ITT and LATE."

plot_sens("duration")
```

## Effect paths at fixed bandwidths

Each family's path across horizons 2–36 months, re-estimated at the family's
fixed bandwidth and overlaid on the paper's CCT-optimal-bandwidth estimates.
With a fixed $h$, every point within a panel uses the same local sample, so
the shape of the path is no longer confounded by bandwidth movement.

```{r}
#| label: fig-path-fns
#| output: false

plot_path <- function(fam, est, ylab) {
  bind_rows(
    paths_fixed |>
      filter(family == fam, estimand == est) |>
      mutate(Bandwidth = "Fixed h"),
    paths_opt |>
      filter(family == fam, estimand == est) |>
      mutate(Bandwidth = "CCT-optimal h")
  ) |>
    ggplot(aes(months, coef, color = Bandwidth)) +
    geom_hline(yintercept = 0, color = "gray60") +
    geom_pointrange(
      aes(ymin = ci_l, ymax = ci_u),
      position = position_dodge(width = 0.8),
      size = 0.3
    ) +
    scale_x_continuous(breaks = seq(3, 36, 3)) +
    scale_color_manual(
      values = c("CCT-optimal h" = "gray50", "Fixed h" = "steelblue")
    ) +
    labs(
      x = "Months since displacement",
      y = ylab,
      title = est,
      color = NULL
    )
}
```

```{r}
#| label: fig-path-takeup
#| fig-cap: "Take-up path (first stage) at fixed vs. CCT-optimal bandwidth."

plot_path("take_up", "First stage", "Take-up rate")
```

```{r}
#| label: fig-path-survival
#| fig-height: 8
#| fig-cap: "Survival paths at fixed vs. CCT-optimal bandwidth, ITT and LATE."

ggarrange(
  plot_path("survival", "ITT", "Survival rate"),
  plot_path("survival", "LATE", "Survival rate"),
  nrow = 2, common.legend = TRUE, legend = "bottom"
)
```

```{r}
#| label: fig-path-duration
#| fig-height: 8
#| fig-cap: "Duration paths at fixed vs. CCT-optimal bandwidth, ITT and LATE."

ggarrange(
  plot_path("duration", "ITT", "Censored duration (weeks)"),
  plot_path("duration", "LATE", "Censored duration (weeks)"),
  nrow = 2, common.legend = TRUE, legend = "bottom"
)
```

## Summary: optimal vs. fixed bandwidth at the anchors

```{r}
#| label: tbl-summary
#| tbl-cap: "Anchor estimates under the CCT-optimal and the fixed bandwidth."

bind_rows(
  anchor_fits |>
    select(family, estimand, h, b, n_eff, coef, se, pv) |>
    mutate(Bandwidth = "CCT-optimal h"),
  paths_fixed |>
    filter(
      (family == "take_up" & months == 12) |
        (family == "survival" & months == 24) |
        (family == "duration" & months == 36)
    ) |>
    select(family, estimand, h, b, n_eff, coef, se, pv) |>
    mutate(Bandwidth = "Fixed h")
) |>
  arrange(family, estimand, Bandwidth) |>
  mutate(family = family_labels[as.character(family)]) |>
  select(
    Family = family, Estimand = estimand, Bandwidth,
    `h (days)` = h, `b (days)` = b, `Effective N` = n_eff,
    Estimate = coef, SE = se, `p-value` = pv
  ) |>
  gt(groupname_col = "Family") |>
  fmt_number(columns = c(`h (days)`, `b (days)`), decimals = 1) |>
  fmt_number(columns = `Effective N`, decimals = 0, use_seps = TRUE) |>
  fmt_number(columns = c(Estimate, SE, `p-value`), decimals = 3) |>
  tab_source_note(
    "Fixed-h rows scale the bias bandwidth b to preserve the anchor's
     rho = b0 / h0. All other specification choices match the paper."
  )
```
