

```{r}
#| include: false
#| cache: false

rm(list = ls())
gc()

pacman::p_load(
  tidyverse,
  here,
  gt,
  gtsummary,
  fixest,
  arrow,
  ggpubr,
  modelsummary,
  knitr,
  kableExtra,
  targets
)

here::i_am("sections/empirical-strategy.qmd")

tar_config_set(store = here("_targets"))

tar_source(files = here("R"))

theme_set(theme_minimal())

```



::: {#fig-density}
```{r}


tar_read(density_plot)

```

*Note:* This figure shows there was no manipulation into eligibility by altering the days of contributions to social security at the moment of displacement [@cattaneo2020].

Density of the Running Variable
:::

::: {#fig-covariates-pre}
```{r}
#| fig-height: 10
#| fig-width: 9

tar_read(covariates_plots)

```

*Note:* This figure shows (*i*) sample means for evenly-spaced 14-day bins relative to the eligibility threshold, and (*ii*) the underlying regression function estimated using a triangular kernel, optimal bandwidth and a linear local polynomial using @calonico2015. The figure shows workers' covariates do not systematically jump at the eligibility threshold.

Covariates around Eligibility Threshold
:::

::: {#fig-pre_unemp}
```{r}
#| fig-height: 8
#| fig-width: 9

tar_read(prev_job_plots) 


```

*Note:* his figure shows (*i*) sample means for evenly-spaced 14-day bins relative to the eligibility threshold, and (*ii*) the underlying regression function estimated using a triangular kernel, optimal bandwidth and a linear local polynomial using @calonico2015. The figure shows workers' previous job outcomes do not systematically jump at the eligibility threshold.

Previous Job Outcomes around Eligibility Threshold
:::

```{r}
#| label: tbl-covariates-bal
#| tbl-cap: "Eligibility Effect: Balance of Covariates"
#| cache: false


tar_read(all_covariates) |>
    arrange(group, label) |>
    slice_head(n = 5) |>
    my_modelsummary(digits = 3) |>
    kable_styling(latex_options = "scale_down") |>
    column_spec(1, width = "11em") |>
    column_spec(2:6, width = "6em")

tar_read(all_covariates) |>
    arrange(group, label) |>
    slice_tail(n = 5) |>
    my_modelsummary(digits = 3) |>
    kable_styling(latex_options = "scale_down") |>
    column_spec(1, width = "11em") |>
    column_spec(2:6, width = "6em") |>
    footnote(
        general = rdd_footnote,
        general_title = "Note:",
        threeparttable = T,
        escape = F,
        footnote_as_chunk = T
    )

```
