

```{r}
#| include: false
#| cache: false

rm(list = ls())
gc()

pacman::p_load(
  tidyverse,
  here,
  gtsummary,
  fixest,
  arrow,
  ggpubr,
  modelsummary,
  knitr,
  tinytable,
  targets
)

here::i_am("sections/empirical-strategy.qmd")

tar_config_set(store = here("_targets"))

tar_source(files = here("R"))

# TeX Gyre Heros is the sansfont in wp.qmd, and is visible to both Typst and R.
# Setting it here is what makes figure text match the captions around it.
theme_set(theme_minimal(base_family = "TeX Gyre Heros"))

```



::: {#fig-density}
```{r}

density_plot <- tar_read(density_plot)


# Display plot
density_plot  


```

*Note:* This figure shows there was no manipulation into eligibility by altering the days of contributions to social security at the moment of displacement [@cattaneo2020].

Density of the Running Variable
:::

::: {#fig-covariates-pre}
```{r}
#| fig-height: 10
#| fig-width: 9

tar_read(covariates_plots) |> arrange_rd_panels(nrow = 3, ncol = 2)

```

*Note:* This figure shows (*i*) sample means for evenly-spaced 14-day bins relative to the eligibility threshold, and (*ii*) the underlying regression function estimated using a triangular kernel, optimal bandwidth and a linear local polynomial using @calonico2015. The figure shows workers' covariates do not systematically jump at the eligibility threshold.

Covariates around Eligibility Threshold
:::

::: {#fig-pre_unemp}
```{r}
#| fig-height: 8
#| fig-width: 9

tar_read(prev_job_plots) |> arrange_rd_panels(nrow = 2, ncol = 2)


```

*Note:* his figure shows (*i*) sample means for evenly-spaced 14-day bins relative to the eligibility threshold, and (*ii*) the underlying regression function estimated using a triangular kernel, optimal bandwidth and a linear local polynomial using @calonico2015. The figure shows workers' previous job outcomes do not systematically jump at the eligibility threshold.

Previous Job Outcomes around Eligibility Threshold
:::

@tbl-covariates-bal reports the formal continuity tests for the ten predetermined covariates. Three of them reject continuity at the 5% level individually, but these p-values are not the right basis for a conclusion: ten simultaneous tests are being conducted, and under the null one would expect roughly one rejection at 5% by chance alone. Applying the Holm step-down correction, which controls the probability of at least one false rejection without assuming independence across covariates, **no covariate retains significance at the 5% level** [@aickin1996]. The correction matters here in particular because two of the three individually significant covariates — age and date of birth — are mechanically the same test with opposite sign, and so are perfectly dependent.

The discontinuities that are individually detectable are also economically negligible. The jump in age amounts to about a month and a half against a sample mean of roughly twenty-five years, and the jump in days since the pension account was opened is on the order of two days against a mean above twelve hundred. Neither is of a magnitude that could plausibly generate the effects reported below.

```{r}
#| label: tbl-covariates-bal
#| tbl-cap: "Continuity of Predetermined Covariates at the Eligibility Threshold"
#| cache: false

# Reads the tidied balance target rather than re-running rdrobust, and adds the Holm
# column: with ten simultaneous tests the raw p-values overstate the evidence against
# continuity, and two of the ten (age and date of birth) are the same test twice.
labs_bal <- c(
    "Mujer"                                = "Female (share)",
    "Fecha de nacimiento"                  = "Date of birth (years)",
    "Inicio de vida laboral formal"        = "Formal labor market entry (years)",
    "Fecha de separación"                  = "Date of separation (years)",
    "Edad"                                 = "Age (years)",
    "Días desde apertura de la cuenta"     = "Days since account opened",
    "Sin CURP"                             = "No CURP (share)",
    "Duración del empleo previo (semanas)" = "Previous job duration (weeks)",
    "Ingresos totales, empleo previo"      = "Previous job total earnings (2024 MXN)",
    "Salario mensual, empleo previo"       = "Previous job monthly wage (2024 MXN)"
)

sig3 <- function(x) {
    vapply(x, function(v) format(signif(v, 3), big.mark = ",", scientific = FALSE),
           character(1))
}

tar_read(balance_table_esp) |>
    mutate(
        covariable = labs_bal[covariable],
        pv_holm    = p.adjust(pv, method = "holm"),
        across(c(coef, se), sig3),
        across(c(pv, pv_holm),
               ~ ifelse(.x < 0.001, "$<$0.001", format(round(.x, 3), nsmall = 3))),
        n_efectivo = format(n_efectivo, big.mark = ","),
        h          = round(h)
    ) |>
    select(covariable, coef, se, pv, pv_holm, n_efectivo, h) |>
    # The `$...$` spans in the column names and in the p-value cells are Typst
    # math, and happen to be spelled the same way they were in LaTeX. tinytable
    # does not escape cell contents, so they pass through as written.
    setNames(c("Covariate", "RD effect", "Std. err.", "$p$",
               "$p$ (Holm)", "Eff. $N$", "$h$ (days)")) |>
    tt(
        width = c(0.28, rep(0.12, 6)),
        notes = my_note(
            "Sharp RD (reduced form) estimates for each predetermined covariate, with a ",
            "local linear polynomial, triangular kernel, optimal bandwidth (Calonico et ",
            "al., 2020) and robust bias-corrected p-values (Calonico et al., 2014). The ",
            "Holm column corrects for the ten simultaneous tests; under that adjustment ",
            "no covariate retains significance at the 5 percent level."
        )
    ) |>
    style_tt(j = 1, align = "l") |>
    style_tt(j = 2:7, align = "r") |>
    style_tt(fontsize = 0.8)

```
